Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-13 paper/sec-mva.tex
Work on paper

After the traditional cut-based analysis presented in the previous section,
and combining the significance of the different event categories,
we end up with a $S/\sqrt{B}$ value around $1.2$.
%
This is not enough to claim observation in this channel.
%
however, the attentive reader might have noticed that we have used
relatively 
loose kinematics cuts, for example in the Higgs mass window.
%
The motivation for this is that the $n$-tuples of signal and background
events, after our basic selection, are processed though a multivariate
analysis with the motivation to increase the signal over background
significance.
%
Multivariate techniques are by now a common tool to enhance signal
significance in HEP analysis, and rightly so,
since they open a new window to improve the performance
of many physics analysis and searches.
%
For instance, the UCL $hh\to 4b$ analysis~\cite{Wardrope:2014kya}
uses a Boosted
Decision Tree MVA to improve signal discrimination.

@@

The classification of events as arising from either signal or
background processes via the use of a multivariate analysis is a common
technique in high energy physics~\cite{Baldi:2014pta,Wardrope:2014kya}.
%
In our work we use a specific type of  MVA to
disentangle signal and background events in the $hh\to 4b$ process,
in particular we use a feed-forward artificial neural network (ANN).
%
The basic idea is that the ANN is given both all MC signal and background
events which satisfy the basic selection cuts {\bf C1}, allowing to identify
which are the most relevant variables to discriminate between the two.

@@

 Therefore, the training of the neural networks consist on the
 minimization of the cross-entropy error,
 Eq.~\ref{cross-entropy}, which in this work is achieved using a
 Genetic Algorithm.
 %
 Since given the large number of events the probability of
 over-fitting is tiny, we just train the MVA for a very long
 number of generations, in this case $N_{\rm gen}=50k$.
 %
 We have verified that if a larger number of generations
 are used the results are unchanged.
 %
 We have also verified that using a cross-validation stopping
 criterion as in the NNPDF3.0 fits does not affect the results.

@@

 To identify the most discriminatory variables, we have to train
 the MVA using a large number of kinematic distributions for
 signal and background events.
%
In our $hh\to 4b$ case,
the input variables for the MVA differ between our three categories,
since in the case of large-$R$ jet reconstruction we want to exploit
the richness of substructure information.

For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA:
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs, $p_{T,h_1}$ and $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$ and $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in $R$ between the two Higgs candidates, $\Delta R_{hh}$.
\item The separation in $\phi$  between the two Higgs candidates, $\Delta \phi_{hh}$.
\item The separation in $\eta$  between the two Higgs candidates, $\Delta \eta_{hh}$.
\item In the boosted category,
  the transverse momenta of the leading, $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and
  sub-leading, $p_{T,h_{2,1}}$ and $p_{T,h_{2,1}}$, Higgs candidate subjets.
  %
  In the resolved category instead, we use the transverse momenta $p_{T,i}$ of the four leading 
   $b$-tagged small-$R$ jets in the event. 
\end{itemize}
Therefore, we have a total of 13 variables which are common to the three categories.
%

For analyses involving the use of large-$R$ jets,
namely for the boosted and intermediate categories,
a number of jet substructure variables~\cite{Aad:2013gja} is used
for every large-$R$ jet that is used for the Higgs reconstruction.
%
In particular we include in the MVA training for each large-$R$ jet
\begin{itemize}
\item The $k_T$ splitting scale~\cite{Butterworth:2002tt,Butterworth:2008iy}.

  This variable is defined by reclustering the constituents of a jet with the
  $k_t$ algorithm, which usually clusters last the harder constituents, and then
  taking the $k_t$ distance measure between the two subjets at the final stage of the recombination
  \be
  \label{eq:ktsplitting}
\sqrt{d_{12}} \equiv {\rm min}\lp p_{T,1},p_{T,2}\rp \cdot \Delta R_{12} \, ,
\ee
where one can also define related variables like $\sqrt{d_{23}}$ that corresponds to the
splitting scale in the second-to-last clustering.

@@

  The $N$-subjettiness variables $\tau_N$ are defined by clustering the constituents
  of a jet with the exclusive $k_t$ algorithm and requiring that $N$ subjets are found,
  \be
  \tau_N \equiv \frac{1}{d_0} \sum_k p_{T,k}\cdot {\rm min}\lp \delta R_{1k}, \ldots,
  \delta R_{Nk}\rp \, , \qquad d_0\equiv \sum_k p_{T,k}\cdot R \, ,
  \ee
  with $ p_{T,k}$ the $p_T$ of the constituent $k$ and $\delta R_{ik}$ the distance from
  subjet $i$ to constituent $k$.
  %
  In this work in particular we use as input to the MVA
  \be
  \label{eq:tau21}
\tau_{21} \equiv \frac{\tau_2}{\tau_1} \, ,
  \ee
  which provide good discrimination between QCD jets and jets arising from the decay of
  a heavy resonance.

\item The ratios of energy correlation functions (EDFs)  $C^{(\beta)}_2$~\cite{Larkoski:2013eya} and
  $D_2^{(\beta)}$~\cite{Larkoski:2014gra}.

  The ratio of energy correlation functions $C_2^{\beta}$ is defined as
  \be
  \label{eq:c2}
C_2^{(\beta)} \equiv \frac{ {\rm ECF}(3,\beta) {\rm ECF}(1,\beta)}{\lc {\rm ECF}(2,\beta)\rc ^2} \, ,
\ee
while $D_2^{(\beta)}$ is instead defined as a double ratio of ECFs, that is
\be
e_3^{(\beta)}\equiv \frac{ {\rm ECF}(3,\beta)}{\lc {\rm ECF}(1,\beta)\rc^3} \, , \quad
  e_2^{(\beta)}\equiv \frac{ {\rm ECF}(2,\beta)}{\lc {\rm ECF}(1,\beta)\rc^2} \, , \quad
  \label{eq:d2}
D_2^{(\beta)} \equiv \frac{ e_3^{(\beta)})}{\lp e_2^{(\beta)} \rp^3} \, .
\ee
The energy correlation functions ${\rm ECF}(N,\beta)$ are defined
  in~\cite{Larkoski:2013eya} with the motivation that $(N+1)$-point correlators
  are sensitive to $N$-prong substructure.
  %
  The free parameter $\beta$ is set to a value of $\beta=2$,
  as recommended by the authors of Refs.~\cite{Larkoski:2013eya,Larkoski:2014gra}.
\end{itemize}
%
Therefore, there are in total $N_{\mathrm{var}}=13$ variables for the resolved analysis,
17 for the semi-resolved, and 21 for the boosted category.
%
Given that the MVA is able to identify the most discriminatory variables
and to ignore those that have smaller effect, it is advantageous to
include as many substructure variables as possible.
%
As we will show, the use of substructure variables in the MVA is the key to increase the signal
significance in the boosted category.

\subsection{Enhancing signal discrimination with MVA}

@@

A very useful property of MVA's such as the one used in this work
is that they provide direct  physical insight about which of the
input variables contribute the most to the signal over background
separation.
%
In the case of ANNs, this can be quantified by computing the sum
of the absolute value of al the weights connected to a given
input neuron $i$, that is
\be
\label{eq:totweight}
\omega^{\rm tot}_i \equiv \sum_{k=1}^5 \Big|\omega^{(2)}_{ki}\Big| \, ,
\qquad i=1,\ldots,N_{\rm var} \, ,
\ee
with $\omega^{(2)}_{ki}$ the value of the weight connecting
the $k$-th neutron of the second layer with the $i$-th neuron of
the first (input) layer.
%
The input variables with a larger value of $\omega^{\rm tot}_i$ will those
be the ones that play a dominant role in enhancing the signal
discrimination using the MVA.

>>>>

From the cut-based analysis,
combining the significance of the three categories,
we end up with a signal significance $S/\sqrt{B}\simeq 1.2(2.3)$
with all backgrounds (only QCD $4b$) taken into account,
which is not enough to be able to claim observation
of Higgs pair production
in this channel, and even less to extract the trilinear Higgs  coupling
$\lambda$.
%
We now process  the $n$-tuples of the signal and background
events that survive the cut-based
analysis by means of a multivariate
analysis, in order increase the signal over background
significance.
%
Multivariate techniques are by now a common tool to enhance signal
significance in HEP analysis, and this is fully justified
since they open a new window to improve the performance
of many physics analysis and searches.
%
For instance, the UCL $hh\to 4b$ analysis~\cite{Wardrope:2014kya}
uses a Boosted
Decision Tree MVA to improve signal discrimination.

In this section first of all we present the specific MVA that we use,
based on feed-forward multi-layer neural networks.
%
We then describe the input variable to the MVA, including the jet substructure
variables, and then present the results of $S/sqrt{B}$ including
the MVA effects.
%
Since we find that jet substructure variables are a crucial input
to the MVA, we end this section showing various kinematic distributions
of these variables.

{\bf add more refs from ATLAS and CMS}

@@

The classification of events as arising from either signal or
background processes via the use of a multivariate analysis is a common
technique in high energy physics~\cite{Baldi:2014pta,Wardrope:2014kya}.
%
In our work we use a specific type of  MVA to
disentangle signal and background events in the $hh\to 4b$ process,
in particular we use a multi-layer feed-forward artificial neural network (ANN),
also known as a {\it perceptron}.\footnote{This type of ANNs are the same
  as those used to parametrize Parton Distribution Functions
in the NNPDF global analyses~\cite{DelDebbio:2004qj,Ball:2008by,Ball:2011mu,Ball:2010de}.}
%
The basic idea is that the ANN is trained on an input composed on
all  signal and background
events which satisfy the
cut-based analysis, allowing to identify in an automated way
which are the most relevant variables to discriminate between the two.

@@

 Therefore, the training of the neural networks consist on the
 minimization of the cross-entropy error,
 Eq.~(\ref{cross-entropy}), which in this work is achieved using a
 Genetic Algorithm.
 %
 Since given the large number of events the probability of
 over-fitting is tiny, we just train the MVA for a very long
 number of generations, in this case $N_{\rm gen}=50k$.
 %
 We have verified that if a larger number of generations
 are used the results are unchanged.
 %
 We have also verified that using a cross-validation stopping
 criterion (in particular, the same one as
 that used in the NNPDF3.0 fit~\cite{Ball:2014uwa})
 does not modify the results.

@@

 To identify the most discriminatory variables, we have to train
 the MVA using a large number of kinematic distributions for
 signal and background events.
%
In our $hh\to 4b$ case,
the input variables for the MVA differ between our three categories,
and in particular in the case of large-$R$ jets we want to exploit
the richness of substructure information.

For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA:
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs, $p_{T,h_1}$ and $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$.
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$ and $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in $R$ between the two Higgs candidates, $\Delta R_{hh}$.
\item The separation in $\phi$  between the two Higgs candidates, $\Delta \phi_{hh}$.
\item The separation in $\eta$  between the two Higgs candidates, $\Delta \eta_{hh}$.
\end{itemize}
In addition, in the boosted category we use
  the transverse momenta of the leading, $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and
  sub-leading, $p_{T,h_{2,1}}$ and $p_{T,h_{2,1}}$, Higgs candidate subjets.
  %
  In the resolved category instead,
  the corresponding variables are
  the transverse momenta $p_{T,i}$ of the four leading 
  $b$-tagged small-$R$ jets in the event.
  %
  In the intermediate category, we use the
  transverse momenta of the subjets
  from the large-$R$ jet $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and the
 transverse momenta $p_{T,i}$ of the two leading 
  $b$-tagged small-$R$ jets.
  %
Therefore, we have a total of 13 variables which are common to the three categories.
%

For analyses involving the use of large-$R$ jets,
namely for the boosted and intermediate categories,
a number of jet substructure variables~\cite{Salam:2009jx,Aad:2013gja} is used
for every large-$R$ jet that is classified as a Higgs candidate.
%
In particular we include in the MVA training for each large-$R$ jet the following
substructure variables:
\begin{itemize}
\item The $k_T$-splitting scale~\cite{Butterworth:2002tt,Butterworth:2008iy}.

  This variable is defined by reclustering the constituents of a jet with the
  $k_t$ algorithm, which usually clusters last the harder constituents, and then
  taking the $k_t$ distance measure between the two subjets at the final stage of the recombination
  \be
  \label{eq:ktsplitting}
\sqrt{d_{12}} \equiv {\rm min}\lp p_{T,1},p_{T,2}\rp \cdot \Delta R_{12} \, .
\ee
It is also possible to define related variables like $\sqrt{d_{23}}$
that corresponds to the
splitting scale in the second-to-last clustering, though we do not use
it here since it is only relevant for three-prong decays.

@@

  The $N$-subjettiness variables $\tau_N$ are defined by clustering the constituents
  of a jet with the exclusive $k_t$ algorithm~\cite{Catani:1993hr}
  and requiring that $N$ subjets are found,
  \be
  \tau_N \equiv \frac{1}{d_0} \sum_k p_{T,k}\cdot {\rm min}\lp \delta R_{1k}, \ldots,
  \delta R_{Nk}\rp \, , \qquad d_0\equiv \sum_k p_{T,k}\cdot R \, ,
  \ee
  with $p_{T,k}$ is the $p_T$ of the constituent particle $k$ and $\delta R_{ik}$ the distance from
  subjet $i$ to constituent $k$.
  %
  In this work we use as input to the MVA the ratio of 2-subjettiness to 1-subjettiness
  \be
  \label{eq:tau21}
\tau_{21} \equiv \frac{\tau_2}{\tau_1} \, ,
  \ee
  which provides good discrimination power
  between QCD jets and jets arising from the decay of
  a heavy resonance.

\item The ratios of energy correlation functions (ECFs)  $C^{(\beta)}_2$~\cite{Larkoski:2013eya} and
  $D_2^{(\beta)}$~\cite{Larkoski:2014gra}.

  The ratio of energy correlation functions $C_2^{\beta}$ is defined as
  \be
  \label{eq:c2}
C_2^{(\beta)} \equiv \frac{ {\rm ECF}(3,\beta) {\rm ECF}(1,\beta)}{\lc {\rm ECF}(2,\beta)\rc ^2} \, ,
\ee
while $D_2^{(\beta)}$ is instead defined as a double ratio of ECFs, that is
\be
e_3^{(\beta)}\equiv \frac{ {\rm ECF}(3,\beta)}{\lc {\rm ECF}(1,\beta)\rc^3} \, , \quad
  e_2^{(\beta)}\equiv \frac{ {\rm ECF}(2,\beta)}{\lc {\rm ECF}(1,\beta)\rc^2} \, , \quad
  \label{eq:d2}
D_2^{(\beta)} \equiv \frac{ e_3^{(\beta)})}{\lp e_2^{(\beta)} \rp^3} \, .
\ee
The energy correlation functions ${\rm ECF}(N,\beta)$ are defined
  in~\cite{Larkoski:2013eya} with the motivation that $(N+1)$-point correlators
  are sensitive to $N$-prong substructure.
  %
  The free parameter $\beta$ is set to a value of $\beta=2$,
  as recommended by the authors of Refs.~\cite{Larkoski:2013eya,Larkoski:2014gra}.
\end{itemize}
%
Therefore, there are in total $N_{\mathrm{var}}=13$ variables for the resolved analysis,
17 for the semi-resolved, and 21 for the boosted category.
%
Given that the MVA is able to identify the most discriminatory variables
and to ignore those that have smaller effect, it is advantageous to
include as many substructure variables as possible.
%

\subsection{Signal significance after the MVA}

@@

A very useful property of MVA's such as the one used in this work
is that they provide direct  physical insight about which of the
input variables contribute the most to the signal over background
separation.
%
In the case of ANNs, this can be quantified by computing the sum
of the absolute value of al the weights connected to a given
input neuron $i$, that is
\be
\label{eq:totweight}
\omega^{\rm tot}_i \equiv \sum_{k=1}^{n^{(2)}} \Big|\omega^{(2)}_{ki}\Big| \, ,
\qquad i=1,\ldots,N_{\rm var} \, ,
\ee
with $\omega^{(2)}_{ki}$ the value of the weight connecting
the $k$-th neutron of the second layer with the $i$-th neuron of
the first (input) layer, and $n^{(2)}=5$ the number of
neurons in the second layer.
%
The input variables with a larger value of $\omega^{\rm tot}_i$ will those
be the ones that play a dominant role in enhancing the signal
discrimination using the MVA.
